---
title: "gnafr guide"
format:
    html:
        toc: true
        toc-depth: 3
execute:
    echo: true
---

## Overview

This guide covers the main `gnafr` workflows:

- creating a DuckDB-backed G-NAF database
- loading G-NAF data
- adding custom addresses
- matching addresses with small and larger examples
- using the main tuning parameters
- exploring other features such as the Shiny app, database sampling, and synthetic address generation

For the setup section below, the code is shown as a reference only.
It is not intended to be executed as part of this document.

## Creating the database

### Create the DuckDB file and schema

```{r}
#| eval: false

# ---- Load package ----
library(gnafr)

# ---- Create or open the DuckDB database ----
con <- gnaf_connect("C:/temp/gnaf.duckdb")

# ---- Create the gnafr tables and indexes ----
gnaf_init(con)

# ---- Inspect table counts ----
gnaf_status(con)
```

### Load a G-NAF Core CSV export

Use `gnaf_load()` when you already have a CSV extract ready to load.

```{r}
#| eval: false

# ---- Load one or more G-NAF Core CSV files ----
gnaf_load(con, "C:/temp/gnaf.qld.csv")

# ---- Or load multiple state files in sequence ----
gnaf_load(con, c(
    "C:/temp/gnaf.qld.csv",
    "C:/temp/gnaf.nsw.csv"
))

# ---- Confirm row counts ----
gnaf_status(con)
```

### Load the raw G-NAF Standard PSV directory

Use `gnaf_load_psv()` when working directly from the Geoscape G-NAF Standard distribution.

```{r}
#| eval: false

# ---- Load directly from the Standard directory ----
gnaf_load_psv(
    con,
    gnaf_dir = "C:/temp/gnaf/G-NAF/G-NAF MAY 2026/Standard"
)

# ---- Confirm row counts after load ----
gnaf_status(con)
```

### Add custom addresses

Custom addresses are stored in `custom_addresses` and are matched alongside G-NAF records.

```{r}
#| eval: false

# ---- Load data.table for convenient input creation ----
library(data.table)

# ---- Define one or more custom addresses ----
custom_addresses <- data.table(
    address_label = "LEVEL 2 123 CUSTOM STREET, BRISBANE QLD 4000",
    flat_type = "LEVEL",
    flat_number = "2",
    number_first = 123L,
    street_name = "CUSTOM",
    street_type = "STREET",
    locality_name = "BRISBANE",
    state = "QLD",
    postcode = 4000L,
    longitude = 153.0234,
    latitude = -27.4698
)

# ---- Insert custom addresses into the database ----
gnaf_add(con, custom_addresses)

# ---- Review table counts ----
gnaf_status(con)
```

## Function usage

The remaining examples assume the database has already been set up and that you have reconnected with:

```{r}
# ---- Reconnect to an existing database ----
library(gnafr)
con <- gnaf_connect("C:/temp/gnaf.duckdb")
```

### Small example with `n = 1`

This is the simplest production-style call: one input, one best match.

```{r}
# ---- Match a single address and return the top result ----
one_result <- gnaf_match(
    con,
    addresses = "10 St James Ct, Tamborine Mountain QLD 4272",
    max_results = 1
)

# ---- Keep the most useful output columns ----
one_result[, .(
    input_raw,
    input_standardised,
    matched,
    match_status,
    total_score,
    address_label,
    longitude,
    latitude
)]
```

### Larger example with several inputs

```{r}
# ---- Define a batch of addresses ----
addresses <- c(
    "unit 110 120 musgrave Road red hill 4000 QLD",
    "U110 1120 musgrave rd red hill 4000",
    "Cambridge on the hill 110/120 musgrave road red hill QLD 4000",
    "18-20 drift cl goldsborough QLD 4865",
    "77 broadwater rd mount gravatt east 4122"
)

# ---- Match all addresses and keep the top two candidates per input ----
batch_results <- gnaf_match(
    con,
    addresses,
    max_results = 2,
    min_score = 40,
    verbose = TRUE
)

# ---- Inspect the main match columns ----
batch_results[, .(
    input_id,
    input_raw,
    input_standardised,
    matched,
    match_status,
    match_rank,
    total_score,
    address_label,
    alias_type,
    source
)]
```

### Different parameters

`gnaf_match()` is intended to be tuned depending on whether you want strict production matching or broader audit-style review.

```{r}
# ---- Strict matching with only the best candidate ----
strict_results <- gnaf_match(
    con,
    addresses,
    max_results = 1,
    min_score = 70
)

# ---- Broader auditing with more candidates and a lower threshold ----
audit_results <- gnaf_match(
    con,
    addresses,
    max_results = 5,
    min_score = 20
)

# ---- Exclude custom addresses from matching ----
gnaf_only_results <- gnaf_match(
    con,
    addresses,
    include_custom = FALSE
)

# ---- Disable locality fallback if you only want postcode and state paths ----
no_fallback_results <- gnaf_match(
    con,
    addresses,
    locality_fallback = FALSE
)

# ---- Trigger fallback sooner for poor initial matches ----
fallback_results <- gnaf_match(
    con,
    addresses,
    fallback_threshold = 85
)

# ---- Change score weights for a number-sensitive workflow ----
weighted_results <- gnaf_match(
    con,
    addresses,
    weights = list(
        postcode = 20,
        suburb = 18,
        street_name = 25,
        street_type = 10,
        number = 20,
        flat = 7
    )
)

# ---- Show progress and summary output ----
verbose_results <- gnaf_match(
    con,
    addresses,
    verbose = TRUE
)
```

## Other features

### Shiny app

The package includes an interactive Shiny geocoder. You can either point it at a DuckDB file or reuse an existing connection.

```{r}
#| eval: false

# ---- Launch the app from a database path ----
gnaf_app(db_path = "C:/temp/gnaf.duckdb")

# ---- Or reuse an existing connection ----
gnaf_app(con = con)
```

The app lets you paste one address per line, run `gnaf_match()`, inspect match scores, and review full-string Jaro-Winkler and Jaccard diagnostics in a `reactable` table.

### Sample the database

Use `sample_gnaf()` to pull a quick sample from the connected database. If there is one user table it returns a `data.table`. If there are multiple tables it returns a named list of `data.table`s.

```{r}
# ---- Sample ten rows per table from the connected database ----
sample_rows <- sample_gnaf(con, n = 10)

# ---- Inspect the object shape ----
class(sample_rows)
names(sample_rows)
```

### Create synthetic test data

`address_perturb_sample()` is useful when you want realistic-but-messy inputs for benchmarking or QA.

```{r}
# ---- Take a sample from the database ----
sample_rows <- sample_gnaf(con, n = 100)

# ---- Choose the G-NAF address table regardless of return shape ----
sample_source <- if (is.list(sample_rows)) sample_rows$gnaf_addresses else sample_rows

# ---- Perturb the sampled labels to create realistic test inputs ----
synthetic_addresses <- address_perturb_sample(
    sample_source,
    n = 25,
    seed = 42,
    max_changes = 2L
)

# ---- Match the simulated strings back against the database ----
synthetic_results <- gnaf_match(
    con,
    synthetic_addresses$simulated_address,
    max_results = 1,
    verbose = TRUE
)

# ---- Compare original and simulated labels ----
synthetic_addresses[, .(
    address_label,
    simulated_address,
    perturbations
)]
```

### Parse addresses without matching

Use `address_parse()` when you want to inspect how free-text addresses are being interpreted before they ever hit the database.

```{r}
# ---- Parse addresses into structured components ----
parsed_addresses <- address_parse(c(
    "unit 110 120 musgrave Road red hill 4000 QLD",
    "Cambridge on the hill 110/120 musgrave road red hill QLD 4000"
))

parsed_addresses
```

### Inspect and rebuild the locality index

The locality index supports the locality-fallback search path in `gnaf_match()`. It is normally maintained automatically, but it can also be rebuilt manually after bulk maintenance work.

```{r}
# ---- Review current table counts ----
gnaf_status(con)

# ---- Rebuild the locality index if needed ----
gnaf_rebuild_locality_index(con)
```

### Close the connection when finished

```{r}
# ---- Disconnect cleanly when your work is done ----
gnaf_disconnect(con)
```