---
title: "gnafr"
subtitle: "One address-matching engine for the whole office"
author: "Kyle Haynes"
date: today
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: true
    transition: slide
    code-line-numbers: false
    footer: "gnafr — Australian address matching against the G-NAF"
    height: 900
    width: 1200
execute:
  echo: true
  warning: false
---

## Agenda {.smaller incremental=false}

1. The problem — address matching is fragmented across the office
2. The tension — some teams need **exactness**, others need **vagueness**
3. Introducing `gnafr`
4. How it works — parse → score → match
5. Live demo
6. Why this fixes fragmentation
7. Other benefits
8. Getting started

# The problem {background-color="#2a6f97"}

## Address matching is fragmented {.smaller}

Across the office, address matching has grown up **independently, team by team**:

- Different scripts, different spreadsheets, different ad-hoc `VLOOKUP`/regex hacks
- Different definitions of "close enough" — nobody agrees on what a match *is*
- No shared source of truth — the same physical address can be standardised three
  different ways in three different systems
- Every team pays the cost of solving parsing and fuzzy-matching **from scratch**
- When something breaks, nobody can explain *why* an address matched or didn't —
  the logic lives in someone's private script

## It's not just duplication — the requirements genuinely differ {.smaller}

::: {.columns}
::: {.column width="50%"}
### Some teams need exactness

- Billing, compliance, legal records
- A wrong match is worse than no match
- False positives are the expensive failure mode
:::

::: {.column width="50%"}
### Some teams need vagueness

- Catchment / service-area analysis
- Spatial joins, coverage estimates
- A missed match is worse than an approximate one
- False negatives are the expensive failure mode
:::
:::

<br>

**Because no single tool served both, teams built their own — and that's the
root cause of the fragmentation.**

# Introducing gnafr {background-color="#2a6f97"}

## One engine, tunable per use case {.smaller}

`gnafr` is an R package that matches messy, free-text addresses against the
**Geocoded National Address File (G-NAF)** — Australia's authoritative address
dataset (~15 million records).

- Takes a **character vector of free-text addresses**, however messy
- Returns the **best matching G-NAF record(s)**, with full structured fields
  and coordinates
- Returns a **transparent 0–100 confidence score**, broken into six components
- Built for **bulk** use — hundreds of thousands of addresses per call, not
  one-at-a-time lookups
- Built for **production reuse** — a single DuckDB file is the shared source
  of truth, queried directly, cached, and reused across R sessions

::: {.fragment}
The key point for today: **strictness is a parameter, not a different tool.**
:::

## How it works — overview

```{mermaid}
%%{init: {'themeVariables': {'fontSize': '20px'}}}%%
flowchart LR
    A[Messy input] --> B[Parse]
    B --> C[Score against<br/>G-NAF candidates]
    C --> D{Confidence<br/>threshold}
    D -->|meets your bar| E[Matched address<br/>+ coordinates]
    D -->|below your bar| F[Unmatched<br/>flagged for review]
```

Same pipeline for every team. The only thing that changes is **where you set
the bar**.

## Step 1 — Parse {.smaller}

`address_parse()` turns a raw string into structured fields: postcode, state,
locality, street name/type, number, flat/unit — using vectorised pattern
matching that covers ~95% of well-formed Australian addresses, with a
fallback parser for the rest.

```{r}
library(gnafr)

address_parse(c(
  "unit 110 120 musgrave Road red hill 4000 QLD",
  "Cambridge on the hill 110/120 musgrave road red hill QLD 4000",
  "18-20 drift cl goldsborough QLD 4865"
))
```

## Step 2 — Score {.smaller}

Every candidate is scored on **six components** that sum to a `total_score`
out of 100:

| Component | Weight | How it's scored |
|---|---|---|
| Street name | 40 | Jaro-Winkler similarity |
| Postcode | 20 | Exact, with partial credit for ±1–3 |
| Suburb / locality | 15 | Jaro-Winkler similarity |
| Street type | 10 | Exact-tier (Rd vs Ct still gets partial credit) |
| Street number | 10 | Exact, or within a ranged G-NAF record |
| Flat / unit | 5 | Exact or absent on both sides |

::: {.fragment}
Nothing is a black box — every match comes back with the full breakdown, so
**any team can see exactly why a score was what it was.**
:::

## Step 3 — Match {.smaller}

`gnaf_match()` tries a sequence of paths, stopping as soon as an input is
satisfied — cheapest and most precise first:

1. **Exact label match** — already-standardised input, instant
2. **Match cache** — seen this input (or an equivalent one) before? Free
3. **Postcode path** — the hot path for the vast majority of inputs
4. **State path** — no postcode parsed at all
5. **Locality fallback** — postcode was wrong, suburb name recovers it
6. **Street-only fallback** *(opt-in)* — brand-new subdivisions not yet
   numbered in G-NAF

All of this runs as **generated SQL inside DuckDB** — no server, one file,
scans millions of rows per query.

# Live demo {background-color="#2a6f97"}

## Same input, same engine, different bar {.smaller}

```{r}
library(data.table)

con <- gnaf_connect(":memory:")
gnaf_init(con)
gnaf_add(con, data.table(
  address_label = "10 MUSGRAVE ROAD, RED HILL QLD 4059",
  number_first = 10L, street_name = "MUSGRAVE", street_type = "ROAD",
  locality_name = "RED HILL", state = "QLD", postcode = 4059L
))

# Real-world messiness: postcode typo'd by one digit
messy <- "10 Musgrave Road, Red Hill QLD 4058"
```

## The "exactness" team {.smaller}

```{r}
gnaf_match(messy, con, min_score = 95, verbose = FALSE)[
  , .(matched, total_score)
]
```

No match returned — the typo isn't good enough for a **compliance-grade**
threshold. Exactly the behaviour that team needs.

## The "vagueness" team {.smaller}

```{r}
gnaf_match(messy, con, min_score = 60, verbose = FALSE)[
  , .(matched, total_score, score_postcode, address_label)
]
```

Same data, same function call — just a lower `min_score` — and the
**catchment analysis** team gets their approximate match, with a visible
score they can sanity-check.

::: {.fragment}
**One package. One shared database. Two legitimate, opposite requirements —
solved by a parameter, not a parallel tool.**
:::

```{r}
#| echo: false
gnaf_disconnect(con)
```

# Why this fixes fragmentation {background-color="#2a6f97"}

## A single, shared source of truth {.smaller}

- One DuckDB file, one parser, one scoring formula — used office-wide instead
  of re-implemented per team
- Every team's threshold and weighting choice is **explicit and reviewable**,
  not buried in a private script
- Custom, not-yet-in-G-NAF addresses (new subdivisions, PO-adjacent site
  codes) are added once, to a shared `custom_addresses` table, and matched
  everywhere — not re-entered by every team that hits them
- Weights can be re-biased per workflow (e.g. toward street number for a
  number-sensitive job) without touching the underlying engine
- When the matching logic improves, **every team benefits at once** instead
  of one script getting fixed while five others stay broken

# Other benefits {background-color="#2a6f97"}

## Built for scale {.smaller}

- **Bulk-first** — one SQL query batches the whole input vector, not a loop
  per address
- **DuckDB backend** — embedded, columnar, no server to run; scans millions
  of rows per query
- **Persistent match cache** — addresses seen before are free the second
  time, across R sessions
- **`data.table` throughout** — large joins and vectorised scoring stay fast
  even outside the database

## Beyond matching {.smaller}

- **Spatial helpers** — join matched addresses to ABS Mesh Blocks / SA2s,
  plot match density as a heatmap, static (`ggplot2`) or interactive
  (`leaflet`)
- **Shiny app** — a paste-addresses-and-go geocoder for non-technical staff,
  built on the exact same `gnaf_match()` engine — no separate logic to drift
  out of sync
- **Synthetic test data** — `address_perturb_sample()` generates
  realistic-but-messy addresses from your own data, for QA and benchmarking
  without hand-labelling a test set

## Production-ready {.smaller}

- Automated test suite (`testthat`) covering the parser, the scorer, and the
  end-to-end matching pipeline
- Full documentation site — theory, quickstart, function reference,
  troubleshooting
- Honest about scope — it's built for **Australian street addresses**; it
  doesn't resolve PO Boxes / GPO Boxes / Locked Bags, and it isn't a general
  GIS toolkit

## Getting started {.smaller}

```r
library(gnafr)

# One shared database file for the office
con <- gnaf_connect("C:/data/gnaf.duckdb")
gnaf_init(con)
gnaf_load(con, "C:/data/GNAF_CORE.csv")

# Bulk match — same call for every team
results <- gnaf_match(
  my_addresses,
  con,
  min_score = 60   # <- the one knob each team sets for themselves
)
```

Full docs: `docs/` in this repo (quickstart, theory, function reference).

## Questions? {.smaller incremental=false}

**gnafr** — one address-matching engine, shared across the office, tuned per
team instead of rebuilt per team.

Kyle Haynes · kyle@kylehaynes.com.au
