# Quick guide

This chapter gets you from zero to a working match in a few minutes, using a tiny
**in-memory** database so you don't need to download G-NAF first. Everything here
actually runs — the output you see below was generated by the code above it, against
the current package source.

When you're ready to match against real Australian addresses, skip to [Building a
good lookup table](04-build-database.qmd).

## Install

```{r}
#| eval: false
# install.packages("devtools")
devtools::install_github("https://github.com/KyleHaynes/gnafr")
```

## A five-minute, no-download example

A real database is built from a multi-million-row G-NAF file (see [Building a good
lookup table](04-build-database.qmd)). To see the whole API working end to end right
now, this section builds a database with three hand-written addresses instead.

```{r}
library(gnafr)
library(data.table)

# ":memory:" creates a database that lives only for this R session.
con <- gnaf_connect(":memory:")
gnaf_init(con)

demo_addresses <- data.table(
  address_label = c(
    "10 MUSGRAVE ROAD, RED HILL QLD 4059",
    "120 MUSGRAVE ROAD, RED HILL QLD 4059",
    "18-20 DRIFT CLOSE, GOLDSBOROUGH QLD 4865"
  ),
  number_first  = c(10L, 120L, 18L),
  number_last   = c(NA_integer_, NA_integer_, 20L),
  street_name   = c("MUSGRAVE", "MUSGRAVE", "DRIFT"),
  street_type   = c("ROAD", "ROAD", "CLOSE"),
  locality_name = c("RED HILL", "RED HILL", "GOLDSBOROUGH"),
  state         = c("QLD", "QLD", "QLD"),
  postcode      = c(4059L, 4059L, 4865L),
  longitude     = c(153.0066, 153.0080, 145.6203),
  latitude      = c(-27.4570, -27.4575, -17.2806)
)

gnaf_add(con, demo_addresses)
gnaf_status(con)
```

Now match a deliberately messy input string against it:

```{r}
result <- gnaf_match(
  "unit 5 18-20 drift cl goldsborough 4865",
  con,
  max_results = 2,
  verbose = FALSE
)

result[, .(input_raw, matched, total_score, address_label,
           score_postcode, score_street_name, score_number)]
```

A perfectly clean input scores 100:

```{r}
gnaf_match("10 Musgrave Road, Red Hill QLD 4059", con, verbose = FALSE)[
  , .(input_raw, matched, total_score, address_label)
]
```

A typo'd postcode (`4058` instead of `4059`) still matches, with the postcode
component scored down rather than the match being rejected — see [near-miss postcode
credit](03-theory.qmd#sec-scoring) for why:

```{r}
gnaf_match("10 Musgrave Road, Red Hill QLD 4058", con, verbose = FALSE)[
  , .(input_raw, matched, total_score, score_postcode, address_label)
]
```

A street **number** that isn't in the database at all, by contrast, is *never*
returned as a low-scoring match — it's filtered out before scoring even runs (see
[the number pre-filter](03-theory.qmd#sec-matching-paths)):

```{r}
gnaf_match("999 Musgrave Road, Red Hill QLD 4059", con, min_score = 0, verbose = FALSE)[
  , .(input_raw, matched, match_status, total_score, address_label)
]
```

Always close the connection when you're done with it (in-memory databases simply
vanish; file-backed ones should still be disconnected cleanly):

```{r}
gnaf_disconnect(con)
```

## The two calls you'll use most

```{r}
#| eval: false
# Connect once per session, reuse the connection across many gnaf_match() calls
con <- gnaf_connect("C:/temp/gnaf.duckdb")

# Match a vector of addresses — this is the function you'll call repeatedly
results <- gnaf_match(addresses, con, max_results = 1, min_score = 60)
```

`gnaf_match()`'s two most useful tuning knobs:

| Argument | Default | Effect |
|---|---|---|
| `max_results` | `1` | Return up to N candidates per input, ranked best first. |
| `min_score` | `60` | Drop candidates scoring below this. Lower it for audit/review workflows; raise it for stricter automated pipelines. |

## Where to next

- New to the package and want to understand *why* it returns what it returns? Read
  [How matching works](03-theory.qmd).
- Have a real G-NAF download ready? Read [Building a good lookup table](04-build-database.qmd).
- Know what you want to call? Jump straight to the [Function reference](05-function-reference.qmd).
