# How matching works

This chapter explains gnafr from the inside: how a raw string becomes structured
components, how those components become a score, and how `gnaf_match()` decides
which database rows are even worth scoring. Every number quoted here was read from
the current source ([`R/parse.R`](../R/parse.R), [`R/score.R`](../R/score.R),
[`R/match.R`](../R/match.R)) and, where practical, verified by actually running it —
look for the executed code blocks.

```{mermaid}
%%{init: {'themeVariables': {
  'fontSize': '12px',
  'nodePadding': 6,
  'rankSpacing': 20,
  'edgeSpacing': 8,
  'diagramPadding': 6
}}}%%
flowchart LR

    A[gnaf_match input]
    B{Exact label<br/>match?}
    C{In match<br/>cache?}
    D[address_parse]

    A --> B
    B -- yes --> Z[Result]
    B -- no --> C
    C -- yes --> Z
    C -- no --> D

    D --> E{Postcode<br/>parsed?}

    subgraph P[DuckDB Matching]
        direction TB
        F[Postcode path]
        G[State path]
    end

    E -- yes --> F
    E -- no --> G

    F --> H{Best score <=<br/>fallback_threshold<br/>or unmatched?}
    G --> H

    H -- no --> Z
    H -- yes & locality parsed --> I[Locality fallback<br/>fuzzy suburb → postcode]

    I --> J{Still unmatched &<br/>street_only_fallback=TRUE?}

    J -- no --> Z
    J -- yes --> K[Street-only fallback]

    K --> Z
```

## Parsing {#sec-parsing}

`address_parse()` turns a raw string into a row of structured fields:
`in_postcode`, `in_state`, `in_locality`, `in_street_name`, `in_street_type`,
`in_street_suffix`, `in_number_first`, `in_number_last`, `in_number_suffix`,
`in_flat_type`, `in_flat_number`, `in_building_name`.

It runs in two tiers:

1. **Vectorised fast paths** (`.parse_vectorized`) — four `stringi`-based regex
   patterns, applied across the *whole* input vector at once, that together cover
   roughly 95%+ of well-formed Australian addresses:
   - **4a — slash notation**: `FLAT/NUMBER STREETNAME` (`"110/120 Musgrave Rd"`)
   - **4b — flat-type prefix**: `UNIT NUMBER NUMBER STREETNAME` (`"Unit 3 120 Musgrave Rd"`)
   - **4b2 — attached single-letter prefix**: `F8 536 STREETNAME` (`F`=Flat, `A`=Apartment, `U`=Unit; any other letter defaults to Unit)
   - **4c — simple form**: `NUMBER[-NUMBER] STREETNAME` (the most common case by far), with an "implied pair" sub-case for the common Australian convention `<unit> <number> <street>` written with no explicit `UNIT` keyword (`"10 120 Musgrave Rd"`)
2. **A row-by-row fallback** (`.parse_single`) for anything the fast patterns don't
   recognise — uncommon structures, fuzzy-misspelt street types, addresses with no
   identifiable street type at all.

Before any of that, every input has its trailing **postcode/state** stripped (in
either order — `"...QLD 4059"` or `"...4059 QLD"`), then its rightmost **street
type** token located.

::: {.callout-tip}
## The comma hint
If the original string has a comma (`"123 St James Rode, Tamborine Mountain..."`),
the word immediately before the *last* comma is treated as a strong hint for the
street type — even fuzzily, via Jaro-Winkler against the known type list. This
exists specifically to catch two failure modes of a plain rightmost-match search:
coincidental abbreviation collisions inside a multi-word street name (`"St"` inside
`"St James Rode"` matching the `STREET` abbreviation and hiding the real, misspelt
type `"Rode"`), and misspelt types that a pure rightmost scan would miss entirely.
:::

### Live example

```{r}
library(gnafr)

address_parse(c(
  "unit 110 120 musgrave Road red hill 4000 QLD",
  "U110 1120 musgrave rd red hill 4000",
  "Cambridge on the hill 110/120 musgrave road red hill QLD 4000",
  "18-20 drift cl goldsborough QLD 4865",
  "F8 536 example parade brisbane 4000"
))
```

### Normalisation (`normalize = TRUE`, the default)

After structural parsing, common abbreviations in `in_locality` and `in_street_name`
are expanded to the form G-NAF actually uses, so they compare equal at scoring time
instead of merely *similar*:

| Pattern | Expands to |
|---|---|
| `MNT`, `MT` | `MOUNT` |
| `ST` (as a whole word, after the street type has already been extracted) | `SAINT` |
| `NTH`, `STH`, and a leading `N`/`S`/`E`/`W` | `NORTH` / `SOUTH` / `EAST` / `WEST` |
| `CK` | `CREEK` |
| `&` | `AND` |
| Ordinal numerals (`1ST`…`20TH`) | written words (`FIRST`…`TWENTIETH`), street name only |

```{r}
address_parse("Mt Gravatt East", normalize = TRUE)[, .(in_locality)]
address_parse("Mt Gravatt East", normalize = FALSE)[, .(in_locality)]
```

This matters because the suburb/street-name score is Jaro-Winkler *similarity*, not
identity — `"MOUNT GRAVATT EAST"` vs `"MT GRAVATT EAST"` would still score well, but
expanding it first means a correctly-typed-but-abbreviated suburb scores a perfect
match instead of a near-miss.

## Scoring {#sec-scoring}

Every (input, candidate) pair is scored on six components that sum to `total_score`
(0–100). The same logic exists twice in the source — once as DuckDB SQL `CASE`
expressions (`.score_sql_exprs()`, used by every matching path so scoring happens
inside the database), and once as equivalent `data.table` vector operations
(`.score_pairs()`, used for the exact-label fast path). **The current default
weights** (`R/score.R`):

```{r}
#| echo: false
w <- list(postcode = 20, suburb = 15, street_name = 40, street_type = 10, number = 10, flat = 5)
data.frame(Component = names(w), Weight = unlist(w)) |> knitr::kable()
```

::: {.callout-important}
These weights changed from an earlier version of the package (which used
postcode 25 / suburb 20 / street\_name 25 / street\_type 10 / number 12 / flat 8) —
street name now carries the most weight by a wide margin. If you've read older notes
or scripts referencing the old weights, they're stale; `.default_match_weights()` is
always the source of truth, and `gnaf_match(..., weights = list(...))` lets you
override them per call (the override must still sum to 100).
:::

You can pass your own `weights` list to bias matching — e.g. toward street number for
a number-sensitive workflow — but note this **disables the match cache for that
call** (cached scores were computed under the default weights, so reusing them under
different weights would silently misreport).

### Postcode (default weight 20)

Exact match scores full weight. Unlike every other component, postcode also has
**near-miss partial credit** — a difference of 1, 2, or 3 scores 70%, 40%, 20% of the
weight respectively; anything further scores 0.

In practice this only ever fires on the **locality fallback** path (@sec-matching-paths):
the primary postcode-path query joins `ON g.postcode = i.in_postcode`, so by
construction every row it scores already has `score_postcode` at full weight. The
partial-credit branches only become reachable once the locality fallback widens the
join to nearby postcodes (±3) or to whatever postcode a fuzzy-matched locality name
points at, regardless of how far off the stated postcode was:

```{r}
con <- gnaf_connect(":memory:")
gnaf_init(con)
gnaf_add(con, data.table::data.table(
  address_label = "10 MUSGRAVE ROAD, RED HILL QLD 4059",
  number_first = 10L, street_name = "MUSGRAVE", street_type = "ROAD",
  locality_name = "RED HILL", state = "QLD", postcode = 4059L
))

# Postcode typo'd by one digit (4058 vs 4059) — the number and street still
# match exactly, so the locality fallback finds it and scores the postcode
# component at 70% of weight (round(20 * 0.7) = 14) instead of rejecting it.
gnaf_match("10 Musgrave Road, Red Hill QLD 4058", con, verbose = FALSE)[
  , .(matched, total_score, score_postcode, address_label)
]
```

### Suburb / locality (default weight 15) and street name (default weight 40)

Both are `round(weight * jaro_winkler_similarity(input, candidate))`, computed by
DuckDB's built-in `jaro_winkler_similarity()` for every matching path except the
exact-label fast path, which uses `stringdist::stringdist(method = "jw", p = 0.1)` in
R — the same metric, just evaluated on whichever side of the SQL/R boundary that path
runs on. Street name carries by far the largest single weight of any component,
reflecting that it's the strongest discriminator between genuinely different
addresses that happen to share a postcode and number.

### Street type (default weight 10)

Not a similarity score — a three-tier exact comparison, because abbreviations are
already canonicalised during parsing and at load time, so by the time scoring runs,
"the same type" should mean *string-equal*, not merely similar:

- Both sides agree (or both are absent) → full weight
- Exactly one side has a type → 50% of weight
- Both sides have a type, but they disagree → 40% of weight

That 40%-for-disagreement floor is deliberate: a wrong street type (`"Avenue"`
typed where G-NAF has `"Drive"`) is a very common, low-stakes user error and
shouldn't fully cancel out an otherwise strong match.

```{r}
gnaf_add(con, data.table::data.table(
  address_label = "13-27 FAIRWAY DRIVE, CLEAR ISLAND WATERS QLD 4226",
  number_first = 13L, number_last = 27L, street_name = "FAIRWAY", street_type = "DRIVE",
  locality_name = "CLEAR ISLAND WATERS", state = "QLD", postcode = 4226L
))

gnaf_match("13 Fairway Avenue, Clear Island Waters QLD 4226", con, min_score = 0, verbose = FALSE)[
  , .(matched, total_score, score_street_type, address_label)
]
```

### Street number (default weight 10)

Three tiers, mirrored identically in SQL and R:

- Exact match to `number_first` → full weight
- Falls within `number_first`–`number_last` (a ranged G-NAF record, e.g. `13-27`) → 70% of weight
- No match → 0

```{r}
gnaf_match("15 Fairway Drive, Clear Island Waters QLD 4226", con, verbose = FALSE)[
  , .(matched, total_score, score_number, address_label)
]
```

A parsed numeric **suffix** (`"190A"`) is handled as its own rule layered on top:
it only earns credit when the *candidate's* `address_label` literally starts with
`"<number><suffix> "` — e.g. input `190A` only matches a G-NAF row whose label starts
with `"190A "`, not merely a row whose `number_first` is `190`. This avoids treating
`190A` and `190B` (almost certainly different physical units) as equivalent just
because they share a numeric prefix.

::: {.callout-warning}
## The number filter is a gate, not a soft score
Unlike every other component, an unmatched street number doesn't just *score* zero —
in most paths, candidates whose number doesn't satisfy the pre-filter (exact, in
range, or the suffix label-prefix rule) are excluded from the SQL join entirely and
never reach scoring at all. An input with a house number that simply isn't in G-NAF
for that street comes back `matched = FALSE`, `match_status = "no_candidate"` — not
a low-scoring row you can raise by lowering `min_score`:

```{r}
gnaf_match("999 Musgrave Road, Red Hill QLD 4059", local({
  c2 <- gnaf_connect(":memory:"); gnaf_init(c2)
  gnaf_add(c2, data.table::data.table(
    address_label = "10 MUSGRAVE ROAD, RED HILL QLD 4059",
    number_first = 10L, street_name = "MUSGRAVE", street_type = "ROAD",
    locality_name = "RED HILL", state = "QLD", postcode = 4059L
  ))
  c2
}), min_score = 0, verbose = FALSE)[, .(matched, match_status, total_score)]
```

The one exception is the **street-only fallback** (@sec-matching-paths), which
exists precisely for this situation — it matches against number-free street records
so an input can still resolve to the right street when its exact number isn't (yet)
in G-NAF.
:::

### Flat / unit (default weight 5)

Binary: full weight if both sides have no flat (or matching flat numbers), 0
otherwise — including when only one side has a flat number. A flat number on the
input that the candidate doesn't have at all is treated as a mismatch, not a "can't
tell" no-op, since a different unit at the same street number is a meaningfully
different physical address.

## Matching paths {#sec-matching-paths}

`gnaf_match()` tries, in order, stopping as soon as an input is satisfied:

### 1. Exact label match

If the uppercased, trimmed input string equals a G-NAF `address_label` exactly, it's
scored and returned immediately — no parsing-driven join needed. Built for
re-processing addresses gnafr (or G-NAF itself) already standardised. Implemented as
a hash join against a registered temporary table rather than a giant SQL `IN (...)`
list, which keeps it fast even for tens of thousands of inputs.

### 2. Match cache

`gnaf_match_cache` stores `(input_standardised → address_detail_pid, scores)` for
every result that previously scored at or above `cache_threshold` (default 95).
Lookups join on `input_standardised` — the same building/flat/number/street/locality
string `gnaf_match()` itself derives from the parsed input — so a future call with a
*differently worded but equivalently parsed* address can still hit the cache. The
cache is **only consulted/written when `weights` are exactly the package defaults**;
non-default weights bypass it entirely for that call (see [Scoring](#sec-scoring)).

### 3. Postcode path

The hot path for the overwhelming majority of inputs: an equi-join on
`g.postcode = i.in_postcode` against the full address table, restricted by the
number pre-filter and a coarse `jaro_winkler_similarity(street_name) >= 0.3` cutoff
*before* scoring — both exist purely to keep the intermediate joined table small,
since the actual scoring formula runs over whatever survives this filter. This is
deliberately a cheap, exact equi-join: near-miss postcodes are not retrieved here at
all (see the callout above) — that's the locality fallback's job.

### 4. State path

For inputs with no parseable postcode at all: the same query shape, joined on
`g.state = i.in_state` instead. Necessarily slower and noisier (a state-wide scan
instead of a postcode-blocked one), so it's only used when there's truly no postcode
to block on.

### 5. Locality fallback

Fires for any input whose best score so far is at or below `fallback_threshold`
(default 90) — including inputs with *no* result yet — provided a locality
(suburb) name was parsed. It discovers alternative postcodes to retry two ways,
unioned together:

- **Fuzzy locality match**: `jaro_winkler_similarity(g.locality_name, parsed_locality) >= 0.85` against `gnaf_locality_index` (a deduplicated `(locality_name, postcode, state)` table, ~3,000 rows for a single state — small enough to scan cheaply). This finds the right postcode regardless of *how* wrong the stated one was.
- **Postcode offset**: the stated postcode ± 1, 2, or 3. Catches near-miss typos whose locality didn't fuzzy-match (itself misspelt, or simply absent from the input).

Why 90 and not, say, 99? A clean correct match typically scores 95+. A
**coincidental** match — same postcode and house number, completely unrelated street
— can still reach the low-to-mid 80s (20 postcode + partial suburb/street credit + a
lucky number match). 90 leaves enough margin to catch those coincidences and let the
locality scan find the real candidate, without re-triggering on matches that are
already correct.

```{r}
gnaf_add(con, data.table::data.table(
  address_label = "77 BROADWATER ROAD, MOUNT GRAVATT EAST QLD 4122",
  number_first = 77L, street_name = "BROADWATER", street_type = "ROAD",
  locality_name = "MOUNT GRAVATT EAST", state = "QLD", postcode = 4122L
))

# Postcode is wildly wrong (9999); the suburb name (after MT -> MOUNT
# normalisation) is enough for the locality fallback to find it.
gnaf_match("77 Broadwater Road, Mt Gravatt East QLD 9999", con, min_score = 0, verbose = FALSE)[
  , .(matched, match_status, total_score, score_postcode, score_suburb, address_label)
]
```

### 6. Street-only fallback (opt-in: `street_only_fallback = TRUE`)

For inputs still unmatched after every other path. Matches only against
`alias_type = 'street_only'` rows — synthetic, number-free address records derived
from G-NAF by [`gnaf_build_street_aliases()`](05-function-reference.qmd#gnaf_build_street_aliases)
(or the official PSV alias tables via
[`gnaf_load_psv()`](05-function-reference.qmd#gnaf_load_psv)). Numbered inputs never
match these rows through the ordinary paths above — the number pre-filter requires
`number_first IS NULL` on the candidate side to match a number-bearing input, and
street-only rows have no number — so this path exists specifically to give those
inputs *something* once everything number-aware has failed. Typical cause: a brand
new subdivision whose house numbers G-NAF hasn't published yet, or an input that
genuinely has no street number.

```{r}
gnaf_disconnect(con)
```

## Why everything runs inside DuckDB

Every matching path registers the parsed inputs as a temporary DuckDB table
(`duckdb::duckdb_register()`), builds one SQL query that joins, scores, ranks
(`ROW_NUMBER() OVER (PARTITION BY input_id ...)`), and filters by `min_score` —
*then* pulls only the surviving rows into R. A few design choices fall directly out
of this:

- **No window-function aggregate before the score filter.** Computing something
  like `COUNT(*) OVER (...)` before filtering on `total_score` would force DuckDB to
  fully materialise the join first, which is exactly the memory blow-up bulk matching
  is built to avoid.
- **The street-name similarity pre-filter (`>= 0.3`) runs in the `WHERE` clause**,
  before scoring — it exists purely to shrink the candidate set, since a 0.3
  Jaro-Winkler floor is cheap to compute and eliminates the bulk of irrelevant rows
  before the (more expensive, in aggregate) full scoring expression runs over them.
- **Registered temp tables, not literal `IN (...)` lists**, for the exact-label and
  cache lookups — `duckdb_register()` lets DuckDB hash-join against an R
  `data.table` directly, which scales far better than a 50,000-item SQL literal.

## The database schema

`gnaf_init()` creates two structurally identical tables — `gnaf_addresses` and
`custom_addresses` — every G-NAF Core column plus the alias/source bookkeeping
columns gnafr adds (`source`, `alias_type`, `alias_principal`, `principal_pid`,
`primary_secondary`, `primary_pid`, `geocode_type`), three supporting structures, and
runs any schema migrations needed for databases created by an older version of
gnafr (every `ALTER TABLE ... ADD COLUMN IF NOT EXISTS` is safe to re-run):

| Object | Purpose |
|---|---|
| `idx_gnaf_pc`, `idx_cust_pc` | Index on `postcode` — supports the state/locality paths' coarser filtering. |
| `idx_gnaf_pcnum`, `idx_cust_pcnum` | Compound index on `(postcode, number_first)` — the index the hot postcode path actually relies on. |
| `idx_gnaf_label`, `idx_cust_label` | Index on `address_label` — backs the exact-label fast path. |
| `gnaf_locality_index` | Deduplicated `(locality_name, postcode, state)` — the small table the locality fallback's fuzzy suburb scan runs against, instead of scanning every address row. |
| `gnaf_match_cache` | `input_standardised → address_detail_pid` plus all six score components and a `cached_at` timestamp. |

`gnaf_init()` is always safe to re-run, including on a database built by an older
version — that's the entire point of the `IF NOT EXISTS` / `ADD COLUMN IF NOT
EXISTS` pattern throughout.

## The alias system

Three independent mechanisms populate non-core rows, all distinguished by the
`alias_type` column:

- **`street_only`** — derived locally by `gnaf_build_street_aliases()` from your own
  loaded `gnaf_addresses`: one row per unique `(street_name, street_type,
  street_suffix, locality_name, state, postcode)` combination, with `number_first`
  left `NULL`. PIDs are an MD5 of the key fields, so re-running without `overwrite =
  TRUE` is a no-op rather than a pile of duplicates.
- **`LOCALITY:SYN` / `STREET:SYN`** — loaded directly from G-NAF's own official alias
  tables by `gnaf_load_psv()`, when working from the raw "Standard" PSV product
  rather than the simplified Core CSV. These catch a *recognised alternative* suburb
  or street name that G-NAF itself records as a synonym — something the locally
  derived `street_only` aliases can't do, since they only ever drop the house number,
  never substitute a different name.
- **`alias_principal` / `principal_pid`** and **`primary_secondary` / `primary_pid`**
  — not alias *rows*, but columns on every row (core or alias) that capture G-NAF's
  own address relationships: whether a record is the canonical `PRINCIPAL` address or
  an `ALIAS` of one, and whether it's the main `PRIMARY` dwelling or a `SECONDARY`
  sub-dwelling/unit, with a pointer back to the related PID in each case.

`gnaf_match()`'s `alias_types` argument restricts which of these are eligible
candidates for a given call — `NA` selects core (non-alias) rows, and the default
`NULL` matches everything regardless of alias type.
