# Other features

Beyond the core parse-score-match pipeline, gnafr includes an interactive geocoder
app, spatial join/plotting helpers, and synthetic address generation for testing.
This chapter covers all three, plus a closer look at the match cache from a
usage (rather than mechanism) angle.

## The Shiny geocoder app

```r
gnaf_app(db_path = "C:/temp/gnaf.duckdb")
# or, reusing an existing connection:
gnaf_app(con = con)
```

`gnaf_app()` launches an interactive geocoder built on the same `gnaf_match()`
engine as everything else in this manual — nothing in the app does its own
matching. You paste one address per line, it runs `gnaf_match()` against your
connected database, and shows results in a `reactable` table.

Two diagnostics are added on top of the usual match columns specifically for manual
review: full-string **Jaro-Winkler** and **Jaccard** similarity between the raw input
and the matched `address_label`, with gradient highlighting so a low-confidence
*lexical* match stands out even when the structured score looks fine (e.g. a match
that's structurally correct — same number, postcode, type — but textually quite
different from the original string, which can be a sign the parse went down an
unexpected path).

Pass `run = FALSE` to get the `shiny.appobj` back without launching it — useful for
embedding the geocoder inside a larger Shiny app, or for testing.

```r
app_obj <- gnaf_app(con = con, run = FALSE)
```

## Spatial helpers

`gnaf_match()` returns `longitude`/`latitude` for every match — the spatial helpers
exist to take that coordinate and join it to *other* geographies (ABS Mesh Blocks,
SA2 regions, custom catchment polygons) or visualise it.

```r
library(gnafr)

# 1. Read a polygon shapefile and see what columns it has
sa2 <- read_shapefile("C:/temp/sa2/SA2_2021_AUST_GDA2020.shp")

# 2. Optionally subset to the area you care about
sa2_qld <- subset_shapefile(sa2, var = "STE_NAME21", values = "Queensland")

# 3. Point-in-polygon join: attach SA2 attributes to your matched addresses
matched <- gnaf_match(addresses, con, max_results = 1)
geo <- spatial_lookup(
  matched[matched == TRUE], sa2_qld,
  lat = "latitude", lon = "longitude",
  return_cols = c("SA2_CODE21", "SA2_NAME21")
)

# 4. Visualise match density against the polygon boundaries
plot_boundaries_heatmap(sa2_qld, points_dt = geo, bins = 200)

# ...or as an interactive leaflet map
plot_boundaries_heatmap(sa2_qld, points_dt = geo, use_leaflet = TRUE)
```

`spatial_lookup()` processes points in chunks (`chunk_size`, default 100,000) to
bound memory use on large batches, and supports `multiple = "all"` for the (rare)
case where your polygons overlap and you want every match, not just the first.

## Synthetic test data

`address_perturb_sample()` generates realistic-but-messy test addresses by sampling
real rows from your loaded data and perturbing them — useful for QA and benchmarking
`gnaf_match()` without hand-labelling a test set.

```{r}
library(gnafr)
library(data.table)

con <- gnaf_connect(":memory:")
gnaf_init(con)
gnaf_add(con, data.table(
  address_label = c(
    "10 MUSGRAVE ROAD, RED HILL QLD 4059",
    "18-20 DRIFT CLOSE, GOLDSBOROUGH QLD 4865"
  ),
  number_first  = c(10L, 18L), number_last = c(NA_integer_, 20L),
  street_name   = c("MUSGRAVE", "DRIFT"),
  street_type   = c("ROAD", "CLOSE"),
  locality_name = c("RED HILL", "GOLDSBOROUGH"),
  state = c("QLD", "QLD"), postcode = c(4059L, 4865L)
))

sample_rows <- sample_gnaf(con, n = 100)
# gnaf_init() always creates four base tables (gnaf_addresses, custom_addresses,
# gnaf_locality_index, gnaf_match_cache), so sample_gnaf() returns a named list —
# pick whichever address table actually has rows in it.
sample_source <- if (is.list(sample_rows) && !is.data.frame(sample_rows)) {
  sample_rows$custom_addresses
} else {
  sample_rows
}

synthetic <- address_perturb_sample(
  sample_source, n = 5, replace = TRUE, seed = 42, max_changes = 2L
)
synthetic[, .(address_label, simulated_address, perturbations)]
```

Round-trip it straight back through `gnaf_match()` to see how well the matcher
recovers the original record from its own perturbed output — a quick, repeatable
regression check after changing scoring weights or parser logic:

```{r}
roundtrip <- gnaf_match(synthetic$simulated_address, con, max_results = 1, verbose = FALSE)
roundtrip[, .(input_raw, matched, total_score, address_label)]

gnaf_disconnect(con)
```

`max_changes` caps how many perturbations are applied per address (typos,
abbreviation swaps, dropped components); `keep_original = TRUE` (the default) keeps
the unperturbed label alongside the perturbed one so you can compare directly.

## The match cache, day to day {#sec-cache-daily}

The mechanics are covered in [How matching works](03-theory.qmd#sec-matching-paths);
day to day, the cache functions you'll actually reach for are:

- **`gnaf_cache_status(con)`** — quick sanity check: how many entries, how old is the
  oldest one. Good first thing to check if a batch job feels slower than expected
  (an empty or stale cache means every input is going through the full pipeline).
- **`gnaf_cache_history(con, by = "day")`** — cache growth over time, plus the
  score distribution of what's been cached. A sudden drop in `avg_score` for a given
  day is worth investigating — it usually means a batch of genuinely harder or
  messier input, not a regression in the matcher.
- **`gnaf_cache_rollback(con, after = ...)`** — when a batch was matched under
  conditions you later realise were wrong (bad input data, a bug since fixed) and you
  want exactly *that* batch re-matched on the next run, without discarding everything
  else the cache has learned.
- **`gnaf_cache_clear(con)`** — the blunt instrument. Reasonable after a major
  scoring or parsing change you don't trust the old cached results under, but
  prefer `gnaf_cache_rollback()` when you can scope it to a time window instead.

Remember the cache only applies when `gnaf_match()` is called with the default
weights (@sec-scoring) — if you're routinely calling with custom weights, the cache
functions above will simply stay empty, which is expected, not a bug.
