# Alias matching in practice

The [theory chapter](03-theory.qmd#sec-matching-paths) explains the alias system from
the source code outward — what `alias_type` values exist, which functions populate
them, and how `gnaf_match()`'s `alias_types` argument and `street_only_fallback` flag
control which of them are eligible candidates. This chapter takes the opposite
approach: one real address, one real database, and every example actually executed
against it.

None of this works without good alias coverage in the first place — see [Building a
good lookup table](04-build-database.qmd), specifically [Step
3](04-build-database.qmd) (`gnaf_build_street_aliases()`, derived locally) and the
[`gnaf_load_psv()` section](04-build-database.qmd) (official G-NAF alias tables, the
richer source). The database used below was built from the full raw PSV product, so
it carries both kinds.

::: {.callout-note}
## This chapter runs against a real, multi-gigabyte G-NAF database
Every other chapter that touches a real database marks those chunks `eval: false`
(see the note on the [front page](index.qmd)) because the file doesn't exist in a
typical build environment. This one is the exception: it's written and rendered
against an actual loaded database at `C:/temp/gnaf.duckdb`, so every table below is
real output, not a synthetic stand-in. The connection is opened `read_only = TRUE`
and every `gnaf_match()` call below passes `cache = FALSE`, purely so the chapter can
be re-rendered without writing to that database — in normal use you'd leave
`cache = TRUE` (the default) and let successful matches populate the cache as usual.
:::

```{r}
library(gnafr)
library(data.table)

con <- gnaf_connect("C:/temp/gnaf.duckdb", read_only = TRUE)
gnaf_status(con) |> reactable::reactable()
```

```{r}
alias_counts <- DBI::dbGetQuery(con, "
  SELECT COALESCE(alias_type, 'core (no alias)') AS alias_type, COUNT(*) AS n
  FROM gnaf_addresses
  GROUP BY alias_type
  ORDER BY n DESC
")
reactable::reactable(
  alias_counts,
  columns = list(n = reactable::colDef(name = "Rows", format = reactable::colFormat(separators = TRUE)))
)
```

Just over a quarter of this database's rows are alias rows, not core G-NAF records —
mostly `LOCALITY:SYN` (a recognised alternative suburb name for the same postcode),
followed by `STREET:SYN` and the locally-derived `street_only`. Ignoring all of that
on every match would be throwing away real coverage for no reason — which is exactly
what the rest of this chapter sets out to show, alongside the cases where ignoring it
on purpose is the right call.

## Meet 6 Parkland Boulevard, Brisbane City

```{r}
parkland <- DBI::dbGetQuery(con, "
  SELECT address_detail_pid, address_label, locality_name, alias_type
  FROM gnaf_addresses
  WHERE street_name = 'PARKLAND' AND street_type = 'BOULEVARD' AND postcode = 4000
    AND (address_label LIKE '6 PARKLAND%' OR address_label LIKE '6A PARKLAND%')
  ORDER BY address_label, alias_type NULLS FIRST
")
reactable::reactable(parkland, defaultPageSize = 10)
```

Two things worth noticing before going any further:

1. Every alias row's `address_detail_pid` is the core PID plus a suffix
   (`GAQLD163236963_LA426562`, etc.) — a stable giveaway that you're looking at an
   alias of a specific underlying record, not an unrelated address.
2. **`6` and `6A` are genuinely separate G-NAF records**, each with its own full set
   of locality aliases. This matters in a minute: a query that doesn't specify a unit
   or letter suffix gives `gnaf_match()` two real, tied candidates to choose between,
   and it has to pick one deterministically. It does — by `address_detail_pid`, not
   by relevance — which is worth knowing before you see "6 Parkland Blvd" resolve to
   `6A` below and wonder if something's wrong. Nothing is; it's a genuine tie, broken
   the same way every time.

## Alias info is useful

### A postcode anchors the match — any recognised suburb name finds it

With the postcode correct, `gnaf_match()` only needs the alias system to resolve the
*suburb* name. Geoscape (and plenty of real-world senders) use `City`, `Brisbane`,
`Central`, and `Spring Hill` interchangeably with `Brisbane City` for this exact
postcode — all four are official `LOCALITY:SYN` aliases for it:

```{r}
inputs <- c(
  "6 Parkland Blvd, Brisbane City QLD 4000",
  "6 Parkland Blvd, City QLD 4000",
  "6 Parkland Blvd, Brisbane QLD 4000",
  "6 Parkland Blvd, Central QLD 4000",
  "6 Parkland Blvd, Spring Hill QLD 4000"
)
res <- gnaf_match(inputs, con, verbose = FALSE, cache = FALSE)
res[, .(input_raw, matched, total_score, score_suburb, address_label, alias_type, locality_name)] |>
  reactable::reactable(defaultPageSize = 10)
```

Every variant scores exactly as well as the "correct" one. That's the alias system
doing its job invisibly — without `LOCALITY:SYN` rows in the candidate pool, four of
these five inputs would be scoring a low suburb-similarity penalty (or worse) for no
real reason, since `"CITY"` and `"BRISBANE CITY"` aren't textually similar even
though they're the same place.

### Without a postcode, alias info is the only thing standing between a match and a wrong guess

The above is a comfortable example because the postcode already pins down the
right area; suburb name is just confirmation. Drop the postcode, and the suburb name
becomes load-bearing — this is where ignoring alias info stops being a neutral
choice and starts actively producing wrong answers:

```{r}
no_postcode <- "6 Parkland Blvd, City QLD"

alias_aware  <- gnaf_match(no_postcode, con, verbose = FALSE, cache = FALSE)
alias_ignored <- gnaf_match(no_postcode, con, verbose = FALSE, cache = FALSE, alias_types = NA)

rbindlist(list(
  cbind(mode = "Default (aliases eligible)", alias_aware[, .(matched, total_score, address_label, alias_type, locality_name)]),
  cbind(mode = "alias_types = NA (core only)", alias_ignored[, .(matched, total_score, address_label, alias_type, locality_name)])
)) |> reactable::reactable(defaultPageSize = 5)
```

Both rows say `matched = TRUE` — that's the dangerous part. There's no error, no
low-confidence flag forcing a second look. With aliases eligible, `"City"` resolves
straight to the right building via its `LOCALITY:SYN` row. With aliases switched
off, there's no core G-NAF locality actually named `"City"` for `gnaf_match()` to
anchor on, so the state-wide scan falls back to whatever scores best on street-name
similarity alone — in this database, a same-ish-sounding street in a *different
postcode, in a different part of the state*, with a still-respectable score. Read
only the `matched` and `total_score` columns and you'd never know.

### A house number G-NAF hasn't caught up to yet

`street_only_fallback` is the other half of the alias system: it matches against
number-free `street_only` rows when every number-aware path comes up empty —
typically a genuinely new subdivision, or (for the purposes of this example) a unit
number that simply doesn't exist:

```{r}
new_unit <- "999 Parkland Blvd, Brisbane City QLD 4000"

without_fallback <- gnaf_match(new_unit, con, verbose = FALSE, cache = FALSE)
with_fallback    <- gnaf_match(new_unit, con, verbose = FALSE, cache = FALSE, street_only_fallback = TRUE)

rbindlist(list(
  cbind(mode = "street_only_fallback = FALSE (default)", without_fallback[, .(matched, match_status, total_score, address_label, alias_type)]),
  cbind(mode = "street_only_fallback = TRUE",            with_fallback[, .(matched, match_status, total_score, address_label, alias_type)])
)) |> reactable::reactable(defaultPageSize = 5)
```

Without it, `match_status` is `no_candidate` — an honest "nothing here", not a wrong
guess. With it on, the number-free `street_only` alias for Parkland Boulevard,
Brisbane City steps in and the input resolves to the right street, the right
postcode, just without unit-level precision. That's the entire point of this
fallback: it's the one path that trades number-level certainty for *something*
rather than nothing, and it only exists because the alias system derived a row for
it in the first place.

## Ignoring alias info on purpose

Both knobs used above to *get* alias matches have an off switch, and there are
legitimate reasons to flip them:

- **`alias_types = NA`** restricts every standard path (postcode, state, locality
  fallback) to core G-NAF rows only — no `LOCALITY:SYN`, `STREET:SYN`, or
  `street_only` candidates. Reach for this when you specifically want the
  *canonical* G-NAF record back for downstream joins (e.g. against another dataset
  keyed on principal `ADDRESS_DETAIL_PID`s) and would rather get `no_candidate` than
  a technically-correct-but-alias-sourced PID you'd have to resolve back yourself.
- **`street_only_fallback = FALSE`**, the default, simply never engages the
  number-free fallback — an unmatched input stays unmatched rather than resolving to
  a street-level-only guess. Reach for this when a wrong number is a hard failure for
  your use case (e.g. dispatch/delivery) rather than something street-level precision
  can paper over.

::: {.callout-warning}
## `alias_types` doesn't gate the street-only fallback
These two switches are independent, not layers of the same dial. Setting
`alias_types = NA` turns off `LOCALITY:SYN`/`STREET:SYN` rows in the main paths, but
`street_only_fallback = TRUE` still matches against `street_only` rows regardless —
in the source, `.match_street_only_duckdb()` hard-codes
`g.alias_type = 'street_only'` and never looks at `alias_types` at all. If you want
to rule out *every* alias type, including the street-only fallback, the fallback's
own argument is the one that needs to stay `FALSE`, not `alias_types`.
:::

```{r}
gnaf_disconnect(con)
```

## Recap

- Alias coverage comes from [building the database well](04-build-database.qmd) in
  the first place — `gnaf_load_psv()` for official `LOCALITY:SYN`/`STREET:SYN` rows,
  `gnaf_build_street_aliases()` for locally-derived `street_only` rows.
- Left on (the default, `alias_types = NULL`), aliases let recognised alternative
  suburb and street names match at full score, and let `street_only_fallback`
  recover inputs G-NAF's core numbers don't (yet) cover.
- Switched off via `alias_types = NA` and `street_only_fallback = FALSE`, matching
  falls back to core G-NAF records only — slightly less forgiving, but exactly what
  you want when a wrong alias-backed guess is worse than an honest `no_candidate`.
