# Troubleshooting & performance

## Working with results

### Extracting the best match per input

```r
best <- results[match_rank == 1]
```

### Filtering by confidence

```r
# Only high-confidence matches for automated processing
high_conf <- results[match_rank == 1 & total_score >= 80]

# Flag low-confidence matches for manual review instead of dropping them
results[, needs_review := total_score < 70]
```

### Finding inputs that didn't match

`gnaf_match()` keeps every input in its output — even unmatched ones — with `NA`
score/address columns and a `match_status` explaining why:

```{r}
library(gnafr)
con <- gnaf_connect(":memory:")
gnaf_init(con)
gnaf_add(con, data.table::data.table(
  address_label = "10 MUSGRAVE ROAD, RED HILL QLD 4059",
  number_first = 10L, street_name = "MUSGRAVE", street_type = "ROAD",
  locality_name = "RED HILL", state = "QLD", postcode = 4059L
))

results <- gnaf_match(
  c("10 Musgrave Road, Red Hill QLD 4059", "1 Nowhere Street, Fakesuburb QLD 9999"),
  con, verbose = FALSE
)
results[, .(input_raw, matched, match_status, total_score)]
```

```{r}
#| echo: false
gnaf_disconnect(con)
```

`match_status` takes one of four values:

| Value | Meaning |
|---|---|
| `matched` | At least one candidate scored at or above `min_score`. |
| `below_min_score` | Candidates existed but none cleared `min_score` — try lowering it, or inspect with `min_score = 0` to see what was found. |
| `insufficient_parse` | The parser couldn't extract a street name, or found none of postcode/state/locality to search by. Check with `address_parse()`. |
| `no_candidate` | Parsing succeeded but no database row satisfied even the coarse pre-filters (most commonly: the street number genuinely isn't in the database — see [the number filter](03-theory.qmd#sec-scoring)). |

### Joining match results back to your input data

```r
dt_in[, input_id := .I]

geo <- results[match_rank == 1, .(input_id, total_score, longitude, latitude,
                                   address_label, address_detail_pid)]

dt_out <- geo[dt_in, on = "input_id"]
```

### Diagnosing a specific low score

```r
parsed <- address_parse("my problem address")
print(parsed)

result <- gnaf_match("my problem address", con, max_results = 5, min_score = 0)
print(result[, .(match_rank, total_score, score_postcode, score_suburb,
                  score_street_name, score_number, address_label)])
```

Common causes of an unexpectedly low (or missing) score, in roughly descending order
of how often they come up:

- **Street number not in the database for that street.** Returns `no_candidate`, not
  a low score — see the callout in [Scoring](03-theory.qmd#sec-scoring). Confirm with
  `address_parse()` that the number parsed as expected, then check
  `DBI::dbGetQuery(con, "SELECT COUNT(*) FROM gnaf_addresses WHERE postcode = ... AND street_name = '...'")`.
- **Postcode not loaded.** If you loaded one state's extract and the address is from
  another, the postcode genuinely isn't in the database:
  ```r
  DBI::dbGetQuery(con, "SELECT COUNT(*) FROM gnaf_addresses WHERE postcode = 4001")
  ```
  A count of `0` means the postcode isn't in your loaded data, not a matching bug.
- **Suburb spelling diverges sharply** (`"MT GRAVATT"` vs `"MOUNT GRAVATT EAST"`).
  Jaro-Winkler handles small differences and the abbreviations `normalize = TRUE`
  expands, but not radical abbreviations or missing qualifying words.
- **Street type not recognised.** If `score_street_type` is consistently the
  40%-of-weight "disagreement" tier instead of full credit, the input's abbreviation
  may not be in the lookup table:
  ```r
  data.table::fread(system.file("extdata", "street_types.csv", package = "gnafr"))
  ```
  Add the missing abbreviation/canonical pair if it's genuinely absent.
- **A long building name is confusing the parser.** Building names before the
  flat/number can occasionally prevent number extraction — check with
  `address_parse()` whether `in_number_first` actually came through.

## Performance

### Typical throughput

| Input size | Estimated time |
|---|---|
| 1,000 | < 1 second |
| 10,000 | 2–5 seconds |
| 100,000 | 15–45 seconds |
| 500,000 | 2–5 minutes (chunking recommended) |

Times assume a laptop with an SSD and a few million G-NAF rows loaded for the
relevant state(s). Results vary with CPU, how concentrated the input postcodes are,
and the proportion of addresses with no parseable postcode.

### What actually drives runtime

- **Parsing.** For inputs that miss every vectorised fast pattern (@sec-parsing),
  `address_parse()` falls back to a row-by-row parser, which is the dominant cost for
  large batches of unusual input. It parallelises trivially since each row is
  independent:
  ```r
  library(parallel)
  n_cores <- detectCores() - 1L
  chunks  <- split(addresses, cut(seq_along(addresses), n_cores, labels = FALSE))
  parsed_list <- mclapply(chunks, address_parse, mc.cores = n_cores)
  parsed <- data.table::rbindlist(parsed_list)
  ```
- **Postcode concentration.** If a huge share of input addresses share one postcode,
  the candidate set for that postcode's join gets large — the
  `(postcode, number_first)` index keeps the join itself cheap, but a very
  postcode-concentrated batch with a wide spread of *house numbers* still touches
  more rows than a postcode-diverse one.
- **The locality fallback firing often.** It only runs for inputs scoring at or below
  `fallback_threshold` (default 90), but if a large fraction of your input has bad
  postcodes, you're paying for two passes (postcode path, then locality fallback)
  instead of one. Worth checking `gnaf_cache_history()` / a sample of
  `match_status` values if a batch feels slow — a high proportion of
  `below_min_score` or unmatched rows usually means the locality fallback is doing a
  lot of work for not much payoff, and the input data quality is the actual problem.
- **Re-establishing the connection per call.** Reuse a single `con` across many
  `gnaf_match()` calls — for Shiny apps or API services, keep it in a global or
  module-level variable rather than reconnecting per request.

### Chunking very large batches

For inputs exceeding ~500k rows or spanning many postcodes, splitting into chunks
caps peak memory:

```r
chunk_size <- 50000L
ids <- seq_len(nrow(dt_in))
chunks <- split(ids, ceiling(ids / chunk_size))

results_list <- lapply(chunks, function(idx) {
  gnaf_match(dt_in$address_string[idx], con, max_results = 1, min_score = 60)
})

results <- data.table::rbindlist(results_list)
```

The [match cache](03-theory.qmd#sec-matching-paths) makes a second run over
previously-seen input close to free, regardless of chunking — high-confidence
results from any earlier chunk (or earlier run) are served from
`gnaf_match_cache` rather than re-matched.

## Common errors

### "File not found" from `gnaf_load()`

Use forward slashes or doubled backslashes — a single backslash is an escape
character in an R string literal:

```r
gnaf_load(con, "C:/temp/gnaf.qld.csv")      # OK
gnaf_load(con, "C:\\temp\\gnaf.qld.csv")    # also OK
gnaf_load(con, "C:\temp\gnaf.qld.csv")      # wrong — \t is a tab escape
```

### DuckDB version errors after upgrading the `duckdb` package

DuckDB database files are tied to the format version of the engine that created
them. If you see an incompatible-version error after upgrading the `duckdb` R
package, the database file needs rebuilding from source data — there's no in-place
upgrade path:

```r
file.remove("C:/temp/gnaf.duckdb")
con <- gnaf_connect("C:/temp/gnaf.duckdb")
gnaf_init(con)
gnaf_load(con, "C:/temp/gnaf.qld.csv")
```

### `gnaf_cache_clear()` / `gnaf_cache_rollback()` error in a script

Both prompt for confirmation by default (`ask = TRUE`), which requires an
interactive session. Pass `ask = FALSE` for scripted or scheduled use — you're
opting out of the safety prompt deliberately, so make sure the call is scoped
correctly first (see [the cache walkthrough](06-other-features.qmd#sec-cache-daily)).

### Weights error: `'weights' must sum to 100`

`gnaf_match(..., weights = list(...))` validates that every required component is
present, numeric, non-negative, and sums to exactly 100 (within floating-point
tolerance). The required names are `postcode`, `suburb`, `street_name`,
`street_type`, `number`, `flat` — see [current default weights](03-theory.qmd#sec-scoring)
for a sensible starting point to adjust from.
